---
title: "Assignment 2: Wrapping up regression"
author: 
  - "Alex Greb"
  - "DS303, SP25"
  - "Prof. Amber Camp"
date: 2/28/25
format: html
toc: true
editor: visual
theme: spacelab
---

# Assignment 2: Wrapping up linear and logistic regression\*

Assignment 2 covers linear and logistic regression models and includes many of the topics we have covered over the entire semester. You will be using `palmerpenguins` data, which includes a nice variety of continuous and categorical predictors.

To submit this assignment, render this file and save as a pdf. Upload the pdf to Canvas.

\*This assignment does not include mixed effects models, but you will be seeing that on another assignment.

## Load Packages

```{r, message = FALSE}
library(lme4)
library(tidyverse)
# install.packages("palmerpenguins") # install if needed
library(palmerpenguins)
```

## Load Data

Load the `penguins` data and examine it below (use `summary()`, etc.)

```{r}
penguins <- penguins

summary(penguins)
```

## Question 1: Describe the data

What data is contained in this data set? Describe at least four variables (excluding `year`), including what they represent and their data type. Lastly, describe whether you think `year` would be a useful predictor in this data.

The data in this set includes: Species, Island, bill length, bill depth, flipper length, and sex.

-   Species represents the species of penguin; Adelie, Chinstrap, and Gentoo

-   island represents the island that which each penguin was observed on; Biscoe, Dream, and Torgerson.

-   Bill and flipper length represent the length of each penguins bill and flippers respectively, in millimeters.

-   sex represents the sex of each penguin observed.

Year would not be a useful predictor in this data because it simply represents when the data was collected and the range of this data set is only 2007 to 2009 which would not be a large enough margin to see significant change in species characterictics.

## Question 2: EDA

Explore your data visually. Create at least two visualizations that show the relationship between `flipper_length_mm` and its potential predictors.

```{r, warning = FALSE}
ggplot(penguins, aes(x = bill_length_mm, y = flipper_length_mm, color = species)) +
  geom_point()+
  labs(title = "Flipper Length v.s. Bill Length", x = "Bill Length (mm)", y = "Flipper Length (mm)")+
  theme_minimal()

ggplot(penguins, aes(x = species, y = flipper_length_mm, fill = species)) +
  geom_boxplot()+
  labs(title = "Distribution of Flipper Length by Species",
       x = "Species",
       y = "Flipper Length (mm)") +
  theme_minimal()

```

## Question 3: Apply a linear regression

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g`. Interpret the slope and intercept.

```{r}
lm_flipper_body <- lm(flipper_length_mm ~ body_mass_g, data = penguins)

summary(lm_flipper_body)

```

Interpret your model output in your own words below. Be sure to include a sentence explaining how `body_mass_g` impacts `flipper_length_mm` and whether or not the effect is significant.

**Answer:** According to this model body mass does have a significant impact on flipper length, as body mass increases so does flipper length.

## Question 4: Apply a multiple linear regression

Fit a linear regression model predicting `flipper_length_mm` from both `body_mass_g` and `bill_length_mm`. Interpret the slopes and intercept.

```{r}
lm_flipper_body_bill <- lm(flipper_length_mm ~ body_mass_g + bill_length_mm, data = penguins)

summary(lm_flipper_body_bill)

```

Similar to Question 3, interpret the model output in your own words here:

**Answer:** Both body mass and bill length are significantly positively correlated with flipper length, meaning when the penguins weigh more and have longer bills, they will also tend to have longer flippers

## Question 5: Include an interaction

Fit a simple linear regression model predicting `flipper_length_mm` from `body_mass_g,` `bill_length_mm`, and the interaction of the two. Interpret the slopes and intercept.

```{r}
interaction <- lm(flipper_length_mm ~ body_mass_g * bill_length_mm, data = penguins)

summary(interaction)

```

Interpret the model output in your own words below. If there was a change in the pattern of significance, try to explain the logic below as well.

**Answer:** This model is showing that body mass and bill length alone don't directly have correlation to flipper length, however the relationship of body mass and flipper length depends on bill length.

## Question 6: Compare models

Compare the models you built in Questions 4 and 5 using `anova()`.

```{r}
anova(lm_flipper_body_bill, interaction)

anova(lm_flipper_body, lm_flipper_body_bill)
```

-   Which is the better model? How do you know?

    -   **Answer:** model 2 appears to be better because it has more significant correlation with the 2 \*'s

-   Is it possible to compare the models from Questions 3 and 5 using the same method? Why or why not?

    -   **Answer:** I don't believe so because the model from question 3 is a simple linear regression, and the model from question 5 is a multi-linear regression and you would not be able to compare two different kinds of models like that.

## Question 7: Categorical predictors

Build a linear model that includes a categorical predictor of your choice. It is fine to stick with dummy coding. Optional: apply a different coding scheme AND interpret the output correctly for +1 extra credit.

```{r}
lm_flipper_species_body <- lm(flipper_length_mm ~ species + body_mass_g, data = penguins)
summary(lm_flipper_species_body)

```

-   What is the reference level of your categorical predictor?

    -   **Answer: the reference level is Adelie because Chinstrap and Gentoo are visible in the summary**

-   What is your interpretation of this model output? Address all coefficients.

    -   **Answer: According to this model; chinstrap penguins have flippers 5.6 mm longer than Adelie on average, Gentoo penguins have flippers 15,68 mm longer than Adelie on average, and for every 1g of body mass, flipper length increases by .0084 mm**

## Question 8: Relevel your categorical variable

Relevel your categorical variable so that a **different** level becomes the reference. Then, run the same model you did in Question 7 and interpret the output.

Relevel:

```{r}
penguins$species <- relevel(penguins$species, ref = "Chinstrap")

```

Apply model from Question 7:

```{r}
relevel <- lm(flipper_length_mm ~ species + body_mass_g, data = penguins)
summary(relevel)

```

-   What is the new reference level of your categorical predictor?

    -   **Answer: The new reference is Chinstrap penguins**

-   What is your interpretation of this new model output? Address all coefficients.

    -   **Answer: Gentoo penguins have a flipper length 10.08 mm longer than Chinstrap on average, Adelie penguins have flipper length 5.597 mm shorter than chinstrap on average. For every 1 g of body mass, flipper length will increase by .0084 mm still.**

## Question 9: Apply a logistic regression

Apply a logistic regression. Include as many predictor variables as you'd like. Remember that your predicted outcome variable needs to be binary (or categorical with two levels).

\
Hint: You could use `sex` or create a binary variable of your own (e.g., Gentoo vs. non-Gentoo) to test your model.

```{r}
penguins_clean <- penguins %>% drop_na(sex)

logistic_model <- glm(sex ~ body_mass_g + flipper_length_mm + bill_length_mm, 
                   data = penguins_clean, 
                   family = binomial)

summary(logistic_model)
```

What are your key takeaways from this model?

**Answer:** The key takeaways for this model are that body mass, bill length, and flipper length are all significant predictors of sex, with males tending to have higher body mass and longer bills, and females tending to have longer flippers

## Question 10: Synthesize the information

Imagine you’re a biologist studying penguin populations. Which predictors do you think are most important to measure or record in the field to predict flipper length? Why?

**Answer:** If i were a biologist and has to decide what was most important to record in the field to predict flipper length i would say body mass and species would be the most important because body mass directly correlates with flipper length, as well as species. with these two you would be able to predict very distinguishable groups in my opinion to be able to predict flipper length.

## Bonus: Stepwise Regression

Perform stepwise regression to find the best model for an outcome of your choice. You will likely encounter an error – fixing that error and explaining your findings will earn you +1 extra credit. Show your work.

```{r}


```

According to this stepwise regression, explain how the final model was selected.

**Answer:** (Write answer here)
